---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
    citation_package: natbib
    template: C:/Users/meglarse/Desktop/wickham-thesis/doc-setup/svm-latex-ms.tex
title: Thesis Title
subtitle: Thesis Subtitle
author: 
- name: Victoria Wickham
  affiliation: Major and Department 
- name: Dr. Megan L. Larsen
  affiliation: Water Sciences Laboratory, University of Nebraska-Lincoln
abstract: "Some really good stuff here with 1-2 sentences from the INTRO, RESULTS, and CONCLUSIONS. This section should be no longer than 300 words."
keywords: "harmful algal blooms, frequency, intensity, Nebraska lakes"
geometry: margin=0.75in
fontsize: 11pt
fontfamily: mathpazo
bibliography: C:/Users/meglarse/Desktop/wickham-thesis/doc-setup/wickham-thesis-bib.bib
biblio-style: apsr
csl: C:/Users/meglarse/Desktop/wickham-thesis/doc-setup/environmental-science-and-technology.csl
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

<Setup the global options for the R chunks in your document>
```{r setup, include=FALSE}

setwd("C:/Users/meglarse/Desktop/wickham-thesis/")

# Install packages for the document

## install.packages("ggplot2")
require(ggplot2)

# Package options

```

<Setup the front matter for your document>
\clearpage
\tableofcontents <creates a table of contents for your document>
\newpage
\listoftables <creates a list of all tables>
\newpage
\listoffigures <creates a list of all figures>
\newpage

<Add the text for each of your sections. Check out the [RMarkdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) and [Markdown cheatsheet](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet#lists) for more information about text formatting.>

#Introduction

We can create citations like this: 

  - To suppress the author's name: @smith2017a had some really great things to say.
  - Or to include the full citation: One of his other articles completely contradicted the first [@smith2017b]
  
<To create a formatted citation, check out this website: http://truben.no/latex/bibtex/)>

# Methods
## Study system and data collection
Describe the lakes and hwo the data was collected

## Data sources

## Analyses and calculations
### Frequency

### Intensity

## Statistics

# Results
<Data cleanup>
<Calculations>
<Data Reduction>

## Summarize result 1 in a single sentence.
The frequency of algal blooms across the state of Nebraska increased by X\% between 2005 - 2015 (\autoref{fig1}).

<Figure 1>
```{r fig1, fig.cap="A descriptive title about the frequency. \\label{fig1}", echo = FALSE}
dat <- data.frame(seq(1,10,0.5), seq(1,10,0.5))
ggplot(data = dat, aes(x = dat[,1], y = dat[,2])) +
  geom_point(size = 2, color = "red") +
  xlab("Axis 1") +
  ylab("Axis 2") +
  theme_bw()

```

## Summarize result 2 in a single sentence.
Some great text about this!

<Figure 2>

```{r fig2, fig.cap="A descriptive title about the intensity."}
```

# Conclusions

# References

